Towards typed repositories of proofs \\[0.6em] \small \textsf{MIPS 2010}

### 2010-07-07 talks/MIPS2010/slides.tex
Texte handout pour le slide de syntaxe
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@144 403e655e-980a-47bd-8c60-1ee5348bb6b7

\begin{frame}{The impact of changes}
  \vspace{0.5em}
  \begin{center}
    \only<1>{\includegraphics[width=0.75\textwidth]{images/coqide00.png}}%
    \only<2>{\includegraphics[width=0.75\textwidth]{images/coqide0.png}}%
    \only<3>{\includegraphics[width=0.75\textwidth]{images/coqide.png}}%
    \only<4>{\includegraphics[width=0.75\textwidth]{images/coqide2.png}}%
    \only<5>{\includegraphics[width=0.75\textwidth]{images/coqide4.png}}%
    \only<6>{\includegraphics[width=0.75\textwidth]{images/coqide5.png}}%
    \only<7>{\includegraphics[width=0.75\textwidth]{images/coqide6.png}}%
    \only<8>{\includegraphics[width=0.75\textwidth]{images/coqide7.png}}%
  \end{center}
  \pause
  \uncover<2->{
    \begin{itemize}
    \item File-based separate compilation
      \pause
    \item Interaction loop with global undo
    \end{itemize}
  }
\end{frame}

@@

\begin{frame}{Methodology}
  \begin{center}
    \only<1-2>{\fbox{\parbox{20em}{\only<2>\alert{version
            management}\\[1em]%
          \fbox{\parbox{18em}{script files\\[1em]%
              \fbox{\parbox{17em}{parsing\\[1em]%
                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
    \only<3>{\fbox{\parbox{20em}{script files\\[1em]%
          \fbox{\parbox{18em}{\alert{version management}\\[1em]%
              \fbox{\parbox{17em}{parsing\\[1em]%
                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
    \only<4->{\fbox{\parbox{20em}{\alt<-5>{script files}{user
            interaction}\\[1em]%
          \fbox{\parbox{18em}{parsing\\[1em]%
              \fbox{\parbox{17em}{\alert{version management
                    \only<7->{+ proof-checking}}\\[1em]%
                  \uncover<4-6>{\fbox{\parbox{14.5em}{proof-checking\\}}}}}}}}}}%
  \end{center}
  \begin{itemize}
  \item<4-> AST representation
  \item<5-> Explicit dependency DAG
  \item<7-> Typing annotations
  \item<8-> Incremental type-checking
  \end{itemize}
\end{frame}

@@

\begin{frame}{A core meta-language for incremental type-checking}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \begin{block}{Expresses}
        \begin{itemize}
        \item (abstract) Syntax
        \item (object-) Logics
        \item Proofs (-terms)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{block}{Features}
        \begin{itemize}
        \item Typing
        \item Incrementality
        \item Dependency
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
  \vspace{2em}
  \begin{center}
    \emph{A kernel for a typed version control system?}
  \end{center}
\end{frame}

@@

\begin{frame}{Inspiration: the \textsf{git} storage model}
  \begin{center}
    \begin{overlayarea}{\textwidth}{12em}
      \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
      \only<3>{\includegraphics[width=0.85\textwidth]{images/git2.pdf}}%
      \only<4>{\includegraphics[width=0.85\textwidth]{images/git3.pdf}}%
      \only<5>{\includegraphics[width=0.85\textwidth]{images/git4.pdf}}%
      \only<6-9>{\includegraphics[width=0.85\textwidth]{images/git5.pdf}}%
      \only<10>{
        \begin{center}
          \vspace{5em}
          {\Large Let's design a \emph{typed} \textsf{git}}
        \end{center}
      }%
    \end{overlayarea}
  \end{center}%
  \begin{itemize}\small %
  \item<7-> ``Content-adressable''%
  \item<8-> Name reflects content%
  \item<9-> Maximal sharing (or hash-consing)%
  \end{itemize}%
\end{frame}

@@

\begin{frame}{A typed repository}
  % exemple
\end{frame}

@@

\begin{frame}{Incremental type-checking, incrementally}%
  \begin{block}{Syntax}%
    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
      |\ *\] \[\]}%
    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
      \alert{t\ t}\ |\ *\] \[\]}%
    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
      *\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
      \alert{(x = a)\cdot t}\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
      |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
  \end{block}
  \begin{block}{Semantics}
      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
        \alert{\Gamma[x=a:t]}}
      \only<5->{\ |\ \Gamma[x=a:t]}\] 
      \vspace{1em}
      \begin{center}
        \begin{overlayarea}{0.8\textwidth}{5em}
          \only<-6>{ $ \Gamma \vdash t : u $ \hfill `` In environment
            $\Gamma$, term $t$ has type $u$ '' } \only<7>{ $ \Gamma
            \vdash t : u \alert{\Rightarrow \Delta} $ \hfill ``
            From \alert{repository} $\Gamma$, term $t$ of type $u$ \alert{leads to
            the new repository $\Delta$}'' }
        \end{overlayarea}
      \end{center}
  \end{block}
\end{frame}

\begin{frame}{Typing rules}
  \begin{block}{Product - Init}
  \begin{mathpar}
    \large
    \infer                      % prod
    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[x:t]\vdash u :
      *\uncover<2->{\Rightarrow\Delta}}
    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
    \and
    \uncover<3->{                               %init
      \infer{ }
      {\Gamma\vdash x : t \uncover<4->{\Rightarrow\Gamma}}
      \quad {\small [x:t]\in\Gamma}
    }
  \end{mathpar}
  \end{block}
\end{frame}

\begin{frame}{Typing rules}
  \begin{block}{Equality binder}
    \begin{mathpar}
      \large
      \infer
      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[x=a:u]\vdash t :
        *\Rightarrow\Delta }
      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \notin\Gamma}
      \and \pause
      \infer
      {\Gamma\vdash t \{y/x\} : * \Rightarrow\Delta}
      {\Gamma\vdash(x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \in\Gamma}
    \end{mathpar}
  \end{block}
  \pause
  \begin{block}{Application}
    \[ \infer
      {\Gamma\vdash a : (y:u)\cdot t \Rightarrow\Delta}
      {\Gamma\vdash a\ x : t \{x/y\} \Rightarrow\Delta}
      \quad [x : u] \in \Gamma
      \]
  \end{block}
\end{frame}

\begin{frame}{Naming}

\end{frame}

@@

\begin{frame}{Conclusion}

\end{frame}

>>>>

\hrule
\begin{frame}{The impact of changes}
  \vspace{0.5em}
  \begin{center}
    \only<1>{\includegraphics[width=0.75\textwidth]{images/coqide00.png}}%
    \only<2>{\includegraphics[width=0.75\textwidth]{images/coqide0.png}}%
    \only<3>{\includegraphics[width=0.75\textwidth]{images/coqide.png}}%
    \only<4>{\includegraphics[width=0.75\textwidth]{images/coqide2.png}}%
    \only<5>{\includegraphics[width=0.75\textwidth]{images/coqide4.png}}%
    \only<6>{\includegraphics[width=0.75\textwidth]{images/coqide5.png}}%
    \only<7>{\includegraphics[width=0.75\textwidth]{images/coqide6.png}}%
    \only<8>{\includegraphics[width=0.75\textwidth]{images/coqide7.png}}%
  \end{center}
  \pause
  \uncover<2->{
    \begin{itemize}
    \item File-based separate compilation
      \pause
    \item Interaction loop with global undo
    \end{itemize}
  }
\end{frame}
\hrule

@@

\hrule
\begin{frame}{Methodology}
  \begin{center}
    \only<1-2>{\fbox{\parbox{20em}{\only<2>\alert{version
            management}\\[1em]%
          \fbox{\parbox{18em}{script files\\[1em]%
              \fbox{\parbox{17em}{parsing\\[1em]%
                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
    \only<3>{\fbox{\parbox{20em}{script files\\[1em]%
          \fbox{\parbox{18em}{\alert{version management}\\[1em]%
              \fbox{\parbox{17em}{parsing\\[1em]%
                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
    \only<4->{\fbox{\parbox{20em}{\alt<-5>{script files}{user
            interaction}\\[1em]%
          \fbox{\parbox{18em}{parsing\\[1em]%
              \fbox{\parbox{17em}{\alert{version management
                    \only<7->{+ proof-checking}}\\[1em]%
                  \uncover<4-6>{\fbox{\parbox{14.5em}{proof-checking\\}}}}}}}}}}%
  \end{center}
  \begin{itemize}
  \item<4-> AST representation
  \item<5-> Explicit dependency DAG
  \item<7-> Typing annotations
  \item<8-> Incremental type-checking
  \end{itemize}
\end{frame}
\hrule

@@

\hrule
\begin{frame}{A core meta-language for incremental type-checking}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \begin{block}{Expresses}
        \begin{itemize}
        \item (abstract) Syntax
        \item (object-) Logics
        \item Proofs (-terms)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{block}{Features}
        \begin{itemize}
        \item Typing
        \item Incrementality
        \item Dependency
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
  \vspace{2em}
  \begin{center}
    \emph{A kernel for a typed version control system?}
  \end{center}
\end{frame}
\hrule

@@

\hrule
\begin{frame}{Inspiration: the \textsf{git} storage model}
  \begin{center}
    \begin{overlayarea}{\textwidth}{12em}
      \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
      \only<3>{\includegraphics[width=0.85\textwidth]{images/git2.pdf}}%
      \only<4>{\includegraphics[width=0.85\textwidth]{images/git3.pdf}}%
      \only<5>{\includegraphics[width=0.85\textwidth]{images/git4.pdf}}%
      \only<6-9>{\includegraphics[width=0.85\textwidth]{images/git5.pdf}}%
      \only<10>{
        \begin{center}
          \vspace{5em}
          {\Large Let's design a \emph{typed} \textsf{git}}
        \end{center}
      }%
    \end{overlayarea}
  \end{center}%
  \begin{itemize}\small %
  \item<7-> ``Content-adressable''%
  \item<8-> Name reflects content%
  \item<9-> Maximal sharing (or hash-consing)%
  \end{itemize}%
\end{frame}
\hrule

@@

\hrule
\begin{frame}{A typed repository}
  % exemple
\end{frame}
\hrule

@@

\hrule
\begin{frame}{Incremental type-checking, incrementally}%
  \begin{block}{Syntax}%
    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
      |\ *\] \[\]}%
    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
      \alert{t\ t}\ |\ *\] \[\]}%
    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
      *\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
      \alert{(x = a)\cdot t}\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
      |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
  \end{block}
  \begin{block}{Typing}
      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
        \alert{\Gamma[x=a:t]}}
      \only<5->{\ |\ \Gamma[x=a:t]}\] 
      \vspace{0em}
      \begin{center}
        \begin{overlayarea}{0.8\textwidth}{5em}
          \only<-6>{ $ \Gamma \vdash t : u $ \hfill `` In environment
            $\Gamma$, term $t$ has type $u$ '' } \only<7>{ $ \Gamma
            \vdash t : u \alert{\Rightarrow \Delta} $ \hfill ``
            From \alert{repository} $\Gamma$, term $t$ of type $u$ \alert{leads to
            the new repository $\Delta$}'' }
        \end{overlayarea}
      \end{center}
  \end{block}
\end{frame}
\hrule

Let's construct a typed language for this. We start from the usual
Pure Type Systems here, and we will modify it gradually to fit our
needs.

First, we want to name all intermediate results, that is we don't want
to allow compound, deep applications. We remove general application
and make it a list of variables. Then, we need a new binder to
remember these flat applications. It will behave like a \textsf{let}
binding, and we add a corresponding entry to the environment. Note
that in terms, the type is not present whereas it is in the
environment.

We will use this syntax in quite an unusual way as we will write and
check only \emph{types}. Terms will be sequences of binders declaring
and defining inter-depentent pieces of syntax, and what finishes them
(* or $a$) does not matter in most cases. For the moment, we remove
the abstraction, and with it all notion of computation.

These simple modifications allow us to see this system as a typed
repository of proofs with a last modification: the usual judgement
becomes ``From repository Γ, term t of type u leads to repository
∆''. The analogy here is: Γ is the current database of versions, $t$
is a working directory with possible redundancies (it's not maximally
shared), and integration of $t$ into $\Gamma$ leads to $\Delta$, which
is maximally shared.

We thread the environment along the typing process and return it as
the result.

\hrule
\begin{frame}{Typing rules}
  \begin{block}{Product - Init}
  \begin{mathpar}
    \large
    \infer                      % prod
    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[x:t]\vdash u :
      *\uncover<2->{\Rightarrow\Delta}}
    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
    \and
    \uncover<3->{               %init
      \infer{ }
      {\Gamma\vdash x : t \uncover<4->{\Rightarrow\Gamma}}
      \quad {\small [x:t]\in\Gamma}
    }
  \end{mathpar}
  \end{block}
\end{frame}
\hrule

\hrule
\begin{frame}{Typing rules}
  \begin{block}{Equality binder}
    \begin{mathpar}
      \large
      \infer
      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[x=a:u]\vdash t :
        *\Rightarrow\Delta }
      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \notin\Gamma}
      \and \pause
      \infer
      {\Gamma\vdash t \{y/x\} : * \Rightarrow\Delta}
      {\Gamma\vdash(x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \in\Gamma}
    \end{mathpar}
  \end{block}
  \pause
  \begin{block}{Application}
    \[ \infer
      {\Gamma\vdash a : (y:u)\cdot t \Rightarrow\Delta}
      {\Gamma\vdash a\ x : t \{x/y\} \Rightarrow\Delta}
      \quad [x : u] \in \Gamma
      \]
  \end{block}
\end{frame}
\hrule

\hrule
\begin{frame}{Naming}

\end{frame}
\hrule

@@

\hrule
\begin{frame}{Further Work}

\end{frame}
\hrule
